---
title: "BiodiversityMetrics"
author: "P Galante, A Paz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Biodiversity metrics
Using multiple species for landscape-level species metrics

## Species Richness
### Load and manage spatial data
Loading the SDM data and ensuring the names match the phylogeny

Data are from: Henao-Díaz, F. et al. (2020). Atlas de la biodiversidad de Colombia. Primates. Instituto de Investigación de Recursos Biológicos Alexander von Humboldt. Bogotá D. C., Colombia. 51 pp.

```{r}
# Load binary maps/SDMs
primRas <- stack(list.files(path = "inst/extdata/DemoData/SDM/colPrimates_binary", pattern = "\\.tif$", full.names = T))
# Because of species' name resolutions in the phylogeny we use here, we need to rename some of the species in this rasterStack
names(primRas) <- gsub("*_veg", "", names(primRas))
# Drop models that are resolved into one group; Cebus_albifrons
primRas <- dropLayer(primRas, c("Cebus_versicolor", "Cebus_leucocephalus", "Cebus_malitiosus", "Cebus_albifrons"))
oldnames <- names(primRas)
namesToChange <- c("Cheracebus_lugens", "Cheracebus_medemi", "Lagothrix_lagothricha", "Leontocebus_fuscus", "Plecturocebus_caquetensis", "Plecturocebus_discolor", "Plecturocebus_ornatus", "Saimiri_cassiquiarensis")
newNames <- c("Callicebus_lugens", "Callicebus_medemi", "Lagothrix_lagotricha", "Leontocebus_fuscicollis", "Callicebus_caquetensis", "Callicebus_discolor", "Callicebus_ornatus", "Saimiri_sciureus")
oldnames[oldnames %in% namesToChange] <- newNames
names(primRas) <- oldnames
```
 
Calculate species richness by summing the raster values 
```{r}
SR <- sum(primRas, na.rm = T)
plot(SR)
```


## Phylogenetic diversity
Phylogenetic diversity can be calculated from binary range-maps, or as here, thresholded SDMs. We combine these maps with phylogenetic trees to calculate this measure of diversity.

It is important to note that species-experts can edit binary SDMs, such as is done in the R package [*maskRangeR*](https://cmerow.github.io/maskRangeR/) to add or remove areas.

Load packages that will be useful for these analyses
```{r}
library(picante)
library(raster)
```
### Load and manage spatial data
Loading the SDM data and ensuring the names match the phylogeny
```{r}
# Load binary maps/SDMs
primRas <- stack(list.files(path = "inst/extdata/DemoData/SDM/colPrimates_binary", pattern = "\\.tif$", full.names = T))
# Because of species' name resolutions in the phylogeny we use here, we need to rename some of the species in this rasterStack
names(primRas) <- gsub("*_veg", "", names(primRas))
# Drop models that are resolved into one group; Cebus_albifrons
primRas <- dropLayer(primRas, c("Cebus_versicolor", "Cebus_leucocephalus", "Cebus_malitiosus", "Cebus_albifrons"))
oldnames <- names(primRas)
namesToChange <- c("Cheracebus_lugens", "Cheracebus_medemi", "Lagothrix_lagothricha", "Leontocebus_fuscus", "Plecturocebus_caquetensis", "Plecturocebus_discolor", "Plecturocebus_ornatus", "Saimiri_cassiquiarensis")
newNames <- c("Callicebus_lugens", "Callicebus_medemi", "Lagothrix_lagotricha", "Leontocebus_fuscicollis", "Callicebus_caquetensis", "Callicebus_discolor", "Callicebus_ornatus", "Saimiri_sciureus")
oldnames[oldnames %in% namesToChange] <- newNames
names(primRas) <- oldnames
```

Here, we are only interested in the areas relating to Colombia. To mask the SDMs by Colombia, we can download spatial data and use that as a mask.
```{r}
# First, download altitude data for Colombia using the raster package
col <- raster::getData(name = "alt", country = "COL")
col <- crop(col, primRas)
primRas <- crop(primRas, col)
```

### Load phylogenetic tree
```{r}
primTree <- read.nexus("inst/extdata/DemoData/phyloTree/output.nex")
```

Convert NA values to 0
```{r}
# set all cells that have NA values to 0
for (i in 1:nlayers(primRas)){
  primRas[[i]][is.na(primRas[[i]])] <- 0
}
```

Convert the rasterStack to a data.frame
```{r}
# convert raster to dataframe
commDat <- as.data.frame(primRas)
# remove rows that are NA
commDat <- na.omit(commDat)
```

Calculate phylogenetic diversity, here using the 500th tree as an example
```{r}
phydiv <- pd(commDat[,2:ncol(commDat)], primTree[[1]], include.root = TRUE)
```


